---
title: "Homework 7  MATH531"
author: "Doug Nychka"
date: "`r Sys.Date()`"
output: pdf_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Getting started 

Loading the AudiA4 data and creating the X matrix for "broken" line
regression. 

```{r}
library( fields) # load fields package
load("AudiA4.rda" )
head( AudiA4)
mileage<- AudiA4$mileage/1000
price<- AudiA4$price/1000

cut<- 30
  X<- cbind( ifelse( mileage<= cut, 1, 0),
             ifelse( mileage<= cut, mileage, 0),
             ifelse( mileage > cut, 1, 0),
             ifelse( mileage > cut, mileage, 0)
             )
  
fit<-   lm( price ~ X - 1)
```
For those students rusty with linear algebra in R here are OLS estimates found "by hand". 

```{r}
betaHat<- solve(t(X)%*%X)%*%t(X)%*%price
# check 
bothBetas<- cbind(betaHat,fit$coefficients  )
print( bothBetas)

XbetaHat<- X%*%betaHat
test.for.zero(XbetaHat, fit$fitted.values )
```

Agreement with R function!


A simple plot of the results. Note that using the fitted.values avoids have to 
code directly the broken line function that was fit.


```{r}
plot( mileage, price, pch=16, col="grey")
lines( mileage, fit$fitted.values, col="orange", lwd=3)
abline( v= 30, col="black", lty=2)
title("AudiA4 data -- broken line fit by OLS")
```

Highlight discontinuity in unconstrained fit. 
```{r}
plot( mileage, price, pch=16, col="grey",
xlim=c( 20,40), ylim=c(15,35))
lines( mileage, fit$fitted.values, col="orange", lwd=3)
abline( v= 30, col="black", lty=2)
```

