---
title: "R Lab 3"
author: "Jared Andreatta"
date: "`r Sys.Date()`"
output: 
  pdf_document:
     toc: false
---

# Libraries
library() is used to load libraries with various tools. Here we use MASS and ISLR2 to get datasets
```{r}
library(MASS)
library(ISLR2)
```

# Simple Linear Regression
Now, we look at SLR models with a single predictor.

```{r}
head(Boston)
```

### Fitting and predicting
We can use the lm() function to fit an OLS estimate for this dataset. We'll onlyy have a single regressor for now.


```{r}
# Fitting model
lm.fit <- lm(medv~lstat, data=Boston)

# Summary of linear model
summary(lm.fit)

# Attributes of fitted linear model
names(lm.fit)
# Coefficients 
coef(lm.fit)
# Confidence Intervals for Coefficient Estimates
confint(lm.fit)

# Use the predict() function to construct confidence and prediction intervals
# Produce confidence intervals
predict(lm.fit, data.frame(lstat=(c(5,10,15))), interval="confidence")
# Produce prediction intervals
predict(lm.fit, data.frame(lstat=(c(5,10,15))), interval="prediction")
```
### Plotting 
```{r}
# Plotting regression line
attach(Boston)
plot(lstat, medv)
# Fitted line
abline(lm.fit)
# Adjust line width
abline(lm.fit, lwd = 3)
# Make line red color
abline(lm.fit, lwd = 3, col = "red")

# Different plotting options for the data points
# Red color
plot(lstat, medv, col = "red")
# Solid dots
plot(lstat, medv, pch = 20)
# Pluses
plot(lstat, medv, pch = "+")
# All pch symbols
plot(1:20, 1:20, pch = 1:20)

# These two commands produce diagnostic plots for the lm
par(mfrow = c(2, 2))
plot(lm.fit)

# Alternative way to examine residuals
plot(predict(lm.fit), residuals(lm.fit)) # Computed regression residuals
plot(predict(lm.fit), rstudent(lm.fit)) # Studentized residuals
plot(hatvalues(lm.fit)) # Plotting hat values
# which.max identifies the maximum value in the vector
which.max(hatvalues(lm.fit))
```
# Multiple Linear Regression

Here, we will add more predictors to the model.

```{r}
# Here we add age along lstat
lm.fit <- lm(medv ~ lstat + age, data = Boston)
summary(lm.fit)

# We can use . when we call lm() to regress on all predictors
lm.fit <- lm(medv ~ ., data = Boston)
summary(lm.fit)
# We can also use - to drop regressors
lm.fit <- lm(medv ~ . -age, data = Boston)
summary(lm.fit)

# We can also access individual components of the summary as below
summary(lm.fit)$sigma
summary(lm.fit)$r.sq

# VIF
# We can use vif() (from car) to compute the variance inflation factors
library(car)
vif(lm.fit)
```


```{r}
system.time(
  for (i in 1:1000000000){
    print(i)
  }
)
```

